MSB1015_Assignment 2

### 2019-10-13 MSB1015_Assignment2_SuzannetenHage.Rmd
Small tweaks based on proofread
* Fixed some typo's
* Added student number
* Added line breaks
* Made code for all plots more tidy
* Removed one wrong comment about the error

---
title: "MSB1015_Assignment 2"
author: "Suzanne ten Hage"
date: "13 october 2019"
output: html_notebook
---

**Project Description**
Chemical properties, such as the boiling point, can be derived from the structure of a chemical compound. In 1947, Harry Wiener already made a correlation model to link structural features to boiling points (ref1). The idea to use mathematical models to predict chemical properties from compound structures has been expaned since then.

@@

**0. Installation**
The project requires several packages. The piece of code in this section checks automatically for missing packages and installs them.

The rJava package requires Java to be installed. The code has been developed using Java version 1.8.0_191. A tutorial on how to install this Java Version on windows can be found here: https://downlinko.com/download-install-jdk-8-windows.html

```{r}
#Required packages for this code. 
packages <- c("WikidataQueryServiceR", 
              "rJava", 
              "rcdk", 
              "stringi",
              "caTools",
              "pls", 
              "Metrics"
              )

@@

**1. Get the data from WikiData.** 
In this section I use the WikidataQueryServiceR package to run a SPARQL query that obtains all simple alkines with known boiling points and smiles from WikiData.

```{r}

@@

```
Description of WikiData numbers:
* P31 = instance of
* P279 = subclass of
* Q41581 = alkane
* P2102 = boiling point
* P233 = canonical smiles

The SPARQL query can easily be changed to obtain different data. For example, one could get the smiles of alkynes instead of alkanes by changing Q41581 into Q159226. Look at https://www.wikidata.org/wiki/Wikidata:Main_Page to find the right numbers.

**2. Get the descriptors of the alkane structures**
In this section, I use four descriptors for all alkanes from CDK. I obtain the descriptors based on the alkane smiles. These descriptors are used later as predictors in the PLS model.

*APolDescriptor:*
This describes the sum of the atomic polarizabilities.

*WienerNumbersDescriptor:*
This gives the Wiener numbers (see ref1). It returns the Wiener Path Number and the Wiener Polarity Number.

*MDEDescriptor:*
Returns the Molecular Distance Edge.

*FragmentComplexityDescriptor:*
Returns the complexity of a system.

```{r}

@@

**3. Data pre-processing**
Before we use the descriptors and boiling points to train a PLS model, some data pre-processing needs to be done.

@@

```{r}

@@

#Define which columns to remove. To include one of these columns as a predictor, remove the column here.  
remove_columns <- c("MDEC.13", "MDEC.14", "MDEC.22", "MDEC.23", "MDEC.24", "MDEC.33", "MDEC.34", "MDEC.44", "MDEO.11", "MDEO.12", "MDEO.22", "MDEN.11", "MDEN.12", "MDEN.13", "MDEN.22", "MDEN.23", "MDEN.33")

@@

```{r}

#Plot the boiling points to check for outliers. 
plot(data_combined$data_wikidata.bp, main = "Boiling point values", ylab = "boiling point (Kelvin)", cex.main = 0.8)

```
None of the values in figure 1 have a very abnormal value. Thus, there are no outliers that need to be removed.

Furthermore, one could expect that, in general, bigger alkanes have a higher boiling point. To test whether this relationship is also present in our dataset we plot the length of the smiles vs. the boiling points.

```{r}

#Plot length of smiles vs. boiling points to check for a correlation
plot(stri_length(data_wikidata$smiles), data_wikidata$bp, main = "Figure 2: Length of smiles vs. boiling point", xlab = "length of smiles", ylab = "boiling point (Kelvin)", cex.main = 0.8)

@@

**4. Null model**
The null model is the most simple model possible; the mean of the data. This model can be used as a reference to see whether the PLS model performs better.

```{r}

@@

#error: 250.68 degrees Kelvin.

@@

**5. Partial Least Squares**
Here, I train the PLS model with the training data, and use the model to predict the boiling points of the test data. If you are unfamiliar with PLS, this is video a good introduction: https://www.youtube.com/watch?v=AxmqUKYeD-U. In the model I use leave-one-out (LOO) validation, because my data set is small.

```{r}

@@

**6. Visualization**
**6a)** In order to identify the model with the best number of components, it is needed to plot the error vs. the number of components.

```{r}

#Plot the error vs. the number of PLS components
plot(error_PLS, main = "Figure 3: Error vs. number of components", xlab= "number of PLS components", ylab = "Error (Kelvin)", cex.main = 0.8)

@@

*Conclusion:* From figure 3 we see can conclude that the error decreases with adding more components untill we have 3 PLS components. From there, adding more components doesn't decrease the error further. Therefore, from here on I use the PLS model with 3 components. The error is much lower than the error of the null model, indicating the PLS model performs better than the null model.

@@

```{r}

#Plot the predicted data vs. the original data. 
plot(test$data_wikidata.bp, predicted_bp[,,3], xlim=c(300,1000), ylim=c(300,1000), main = "Figure 4: Predicted boiling points vs. measured boiling points", cex.main = 0.8, xlab = "measured boiling points (Kelvin)", ylab = "predicted boiling points (Kelvin)")
lines(test$data_wikidata.bp, rep(null_model, length(test$data_wikidata.bp)), col = "red")
lines(test$data_wikidata.bp, test$data_wikidata.bp, col = "green")
legend("topleft", c("Null model", "Perfect model", "PLS model"), col = c("red", "green", "black"), pch = c(16), bty = 'n')

@@

*Conclusion*: In figure 4, we see that the PLS model (in black) performs much better in predicting the boiling points than the null model (in red). The PLS model comes quite close to the perfect prediction model, shown in green.

**7. References**
ref1: Wiener H. Structural Determination of Paraffin Boiling Points. Journal of the American Chemical Society. 1947 Jan;69(1):17–20.
ref2: https://www.wikidata.org/wiki/Wikidata:Main_Page (12-10-2019)
ref3: Willighagen et al. The Chemistry Development Kit (CDK) v2.0: atom typing, depiction, molecular formulas, and substructure searching. J. Cheminform. 2017; 9(3), doi:10.1186/s13321-017-0220-4
ref4: May and Steinbeck. Efficient ring perception for the Chemistry Development Kit. J. Cheminform. 2014, doi:10.1186/1758-2946-6-3
ref5: Steinbeck et al. Recent Developments of the Chemistry Development Kit (CDK) - An Open-Source Java Library for Chemo- and Bioinformatics. Curr. Pharm. Des. 2006; 12(17):2111-2120, doi:10.2174/138161206777585274
ref6: Steinbeck et al. The Chemistry Development Kit (CDK): An Open-Source Java Library for Chemo- and Bioinformatics. J. Chem. Inf. Comput. Sci. 2003 Mar-Apr; 43(2):493-500, doi:10.1021/ci025584y

>>>>

---
title: "MSB1015_Assignment 2"
author: "Suzanne ten Hage - i6201596"
date: "13 October 2019"
output: html_notebook
---

**Project Description** <br/>
Chemical properties, such as the boiling point, can be derived from the structure of a chemical compound. In 1947, Harry Wiener already made a correlation model to link structural features to boiling points (ref1). The idea to use mathematical models to predict chemical properties from compound structures has been expaned since then.

@@

**0. Installation** <br/>
The project requires several packages. The piece of code in this section checks automatically for missing packages and installs them. The rJava package requires Java to be installed. The code has been developed using Java version 1.8.0_191. A tutorial on how to install this Java Version on windows can be found [here](https://downlinko.com/download-install-jdk-8-windows.html).

```{r include = TRUE}
#Required packages for this code. 
packages <- c("WikidataQueryServiceR", 
              "rJava", 
              "rcdk", 
              "stringi",
              "caTools",
              "pls", 
              "Metrics"
              )

@@

```
**1. Get the data from WikiData.** <br/>
In this section I use the WikidataQueryServiceR package to run a SPARQL query that obtains all simple alkines with known boiling points and smiles from WikiData.

```{r include = TRUE}

@@

Description of WikiData numbers: <br/>
* P31 = instance of <br/>
* P279 = subclass of <br/>
* Q41581 = alkane <br/>
* P2102 = boiling point <br/>
* P233 = canonical smiles <br/>

The SPARQL query can easily be changed to obtain different data. For example, one could get the smiles of alkynes instead of alkanes by changing Q41581 into Q159226. The numbers can be found [here](https://www.wikidata.org/wiki/Wikidata:Main_Page).

**2. Get the descriptors of the alkane structures** <br/>
In this section, I use four descriptors for all alkanes from CDK. I obtain the descriptors based on the alkane smiles. These descriptors are used later as predictors in the PLS model.

*APolDescriptor:* <br/>
This describes the sum of the atomic polarizabilities.

*WienerNumbersDescriptor:* <br/>
This gives the Wiener numbers (see ref1). It returns the Wiener Path Number and the Wiener Polarity Number.

*MDEDescriptor:* <br/>
Returns the Molecular Distance Edge.

*FragmentComplexityDescriptor:* <br/>
Returns the complexity of a system.

```{r include = TRUE}

@@

**3. Data pre-processing** <br/>
Before we use the descriptors and boiling points to train a PLS model, some data pre-processing needs to be done.

@@

```{r include = TRUE}

@@

#Define which columns to remove. To include one of these columns as a predictor, remove the column here.  
remove_columns <- c("MDEC.13", "MDEC.14", "MDEC.22", "MDEC.23", 
                     "MDEC.24", "MDEC.33", "MDEC.34", "MDEC.44", 
                     "MDEO.11", "MDEO.12", "MDEO.22", "MDEN.11", 
                     "MDEN.12", "MDEN.13", "MDEN.22", "MDEN.23", 
                     "MDEN.33"
                    )

@@

```{r include = TRUE}

#Plot the boiling points to check for outliers. 
plot(data_combined$data_wikidata.bp, 
      main = "Boiling point values", 
      ylab = "boiling point (Kelvin)", 
      cex.main = 0.8
     )

None of the values in figure 1 have a very abnormal value. Thus, there are no outliers that need to be removed. Furthermore, one could expect that, in general, bigger alkanes have a higher boiling point. To test whether this relationship is also present in our dataset we plot the length of the smiles vs. the boiling points.

```{r include = TRUE}

#Plot length of smiles vs. boiling points to check for a correlation
plot(stri_length(data_wikidata$smiles), 
      data_wikidata$bp, 
      main = "Figure 2: Length of smiles vs. boiling point", 
      xlab = "length of smiles", 
      ylab = "boiling point (Kelvin)", 
      cex.main = 0.8
     )

@@

**4. Null model** <br/>
The null model is the most simple model possible; the mean of the data. This model can be used as a reference to see whether the PLS model performs better.

```{r include = TRUE}

@@

@@

**5. Partial Least Squares** <br/>
Here, I train the PLS model with the training data, and use the model to predict the boiling points of the test data. If you are unfamiliar with PLS, this is [video](https://www.youtube.com/watch?v=AxmqUKYeD-U) a good introduction. In the model I use leave-one-out (LOO) validation, because my data set is small.

```{r include = TRUE}

@@

**6. Results** <br/>
**6a)** In order to identify the model with the best number of components, it is needed to plot the error vs. the number of components.

```{r include = TRUE}

#Plot the error vs. the number of PLS components
plot(error_PLS, 
      main = "Figure 3: Error vs. number of components", 
      xlab= "number of PLS components", 
      ylab = "Error (Kelvin)", 
      cex.main = 0.8
     )

@@

*Conclusion:* <br/> 
From figure 3 we see can conclude that the error decreases with adding more components untill we have 3 PLS components. From there, adding more components doesn't decrease the error further. Therefore, from here on I use the PLS model with 3 components. The error is much lower than the error of the null model, indicating the PLS model performs better than the null model.

@@

```{r include = TRUE, message = FALSE, warning = FALSE}

#Plot the predicted data vs. the original data. 
plot(test$data_wikidata.bp, 
      predicted_bp[,,3], 
      xlim=c(300,1000), 
      ylim=c(300,1000), 
      main = "Figure 4: Predicted boiling points vs. measured boiling points", 
      cex.main = 0.8, 
      xlab = "measured boiling points (Kelvin)", 
      ylab = "predicted boiling points (Kelvin)"
     )
#Add null model to the plot
lines(test$data_wikidata.bp, 
       rep(null_model, 
           length(test$data_wikidata.bp)
          ), 
       col = "red"
      )
#Add the perfect prediction to the plot
lines(test$data_wikidata.bp, 
       test$data_wikidata.bp, 
       col = "green"
      )
#Add legend
legend("topleft", 
       c("Null model", "Perfect model", "PLS model"), 
       col = c("red", "green", "black"), 
       pch = c(16), 
       bty = 'n'
      )

@@

*Conclusion:* <br/>
In figure 4, we see that the PLS model (in black) performs much better in predicting the boiling points than the null model (in red). The PLS model comes quite close to the perfect prediction model, shown in green.

**7. References** <br/>
ref1: Wiener H. Structural Determination of Paraffin Boiling Points. Journal of the American Chemical Society. 1947 Jan;69(1):17–20. <br/>
ref2: https://www.wikidata.org/wiki/Wikidata:Main_Page (12-10-2019) <br/>
ref3: Willighagen et al. The Chemistry Development Kit (CDK) v2.0: atom typing, depiction, molecular formulas, and substructure searching. J. Cheminform. 2017; 9(3), doi:10.1186/s13321-017-0220-4 <br/>
ref4: May and Steinbeck. Efficient ring perception for the Chemistry Development Kit. J. Cheminform. 2014, doi:10.1186/1758-2946-6-3 <br/>
ref5: Steinbeck et al. Recent Developments of the Chemistry Development Kit (CDK) - An Open-Source Java Library for Chemo- and Bioinformatics. Curr. Pharm. Des. 2006; 12(17):2111-2120, doi:10.2174/138161206777585274 <br/>
ref6: Steinbeck et al. The Chemistry Development Kit (CDK): An Open-Source Java Library for Chemo- and Bioinformatics. J. Chem. Inf. Comput. Sci. 2003 Mar-Apr; 43(2):493-500, doi:10.1021/ci025584y <br/>

*Package references* <br/>
* [WikidataQuerySerivceR](https://CRAN.R-project.org/package=WikidataQueryServiceR) <br/>
  Mikhail Popov (2017). WikidataQueryServiceR: API Client Library for
  'Wikidata Query Service'. R package version 0.1.1. <br/>

*[rJava](https://CRAN.R-project.org/package=rJava) <br/>
  Simon Urbanek (2019). rJava: Low-Level R to Java Interface. R package
  version 0.9-11.  <br/>

*[rcdk](https://cran.r-project.org/web/packages/rcdk/index.html) <br/>
  Guha, R. (2007). 'Chemical Informatics Functionality in R'. Journal
  of Statistical Software 6(18) <br/>

*[stringi](http://www.gagolewski.com/software/stringi/.) <br/>
  Gagolewski M. and others (2019). R package stringi: Character string
  processing facilities.

*[caTools](https://CRAN.R-project.org/package=caTools) <br/>
  Jarek Tuszynski (2019). caTools: Tools: moving window statistics,
  GIF, Base64, ROC AUC, etc.. R package version 1.17.1.2 <br/>

*[pls](https://CRAN.R-project.org/package=pls) <br/>
  Bjørn-Helge Mevik, Ron Wehrens and Kristian Hovde Liland (2019). pls:
  Partial Least Squares and Principal Component Regression. R package
  version 2.7-1.  <br/>

*[Metrics](https://CRAN.R-project.org/package=Metrics) <br/>
  Ben Hamner and Michael Frasco (2018). Metrics: Evaluation Metrics for
  Machine Learning. R package version 0.1.4. <br/>
